Intentional release of native species undermines ecological stability

### 2022-10-03 rmd/manuscript.Rmd
Update manuscript.Rmd

Human demands for natural resources are ever-increasing, such that active interventions are critical to the sustainable management of fisheries, forestry, and wildlife [@laikreCompromisingGeneticDiversity2010]. Releases of captive-bred native species ("intentional release") is a form of the efforts to enhance wild populations of diverse taxa [@snyderLimitationsCaptiveBreeding1996; @laikreCompromisingGeneticDiversity2010]. The magnitude of intentional release for commercial and recreational species is massive; typically, a single program releases tens of thousands to millions of plant, invertebrate (bivalves @haagNorthAmericanFreshwater2012, insects), and vertebrate individuals (fish, birds) into the wild annually (reviewed in [@laikreCompromisingGeneticDiversity2010; @kitadaEconomicEcologicalGenetic2018]). There is growing awareness that intentional release entails ecological risks, including the accumulation of deleterious alleles [@arakiGeneticEffectsCaptive2007], the spread of diseases and parasites in wild populations [@krkosekDecliningWildSalmon2007], and the intensification of intraspecific competition within the enhanced species. Yet, this method is still pervasive in natural resource management likely because of the apparent economic benefit of release programs. For example, Kitada @kitadaEconomicEcologicalGenetic2018 analyzed economic performance of release programs for 12 major fishery resources -- the gross monetary yield exceeded the per-seed cost in nine species, suggesting that release programs are economically profitable in general.

Current debates, however, overlook the fact that we have rarely assessed the community-wide impact of intentional release. Species are all embedded in the complex web of interacting organisms, and the stable coexistence of competing species through density-dependent feedback underpins the emergent stability of ecological communities [@tilmanEcologicalConsequencesChanges1999; @carrollNicheFitnessDifferences2011a; @loreauDoesFunctionalRedundancy2004]. Intentional release may disrupt the sensitive balance of species interactions because it introduces unnaturally high numbers of individuals into the wild [@careyNativeInvadersChallenges2012]. Hence, this form of species management may intervene in the ecological process that allows competing species to coexist, ultimately degrading the long-term community stability. In particular, the community-wide impact must be serious in release programs for fishery resouces as many of them release millions of hatchery individuals into a single water body (e.g., a watershed) every year, which is likely far beyond the environmental capacity of the recipient ecosystem. Evidence for this hypothesis is lacking, however.

@@

Prior to our full-community analysis, we analyzed a two species community to understand the model behavior. The prior analysis revealed distinct parameter spaces predicting destabilizing, neutral, and stabilizing effects of intentional release on community dynamics. In general, intentional release had a neutral to destabilizing effect in a community with slow-growing species ($r_i < 1.5$). Meanwhile, for higher values of $r_i$ ($\ge 1.5$), intentional release stabilized community dynamics by dampening limit cycles or chaos. Carrying capacity and competition also played an important role in shaping the parameter spaces with different effects of intentional release. According to this result, we chose `r nrow(list_param[[2]])` simulation scenarios, differing in the intrinsic growth of an enhanced species $r_1$, competition $\bar{\alpha}$, carrying capacity $K$, and relative fitness $f_R$ (see **Methods**). Under each parameter scenario, we ran `r attr(list_param[[2]], "n_sim")` time steps of `r max(list_sim[[2]]$n_rep)` independent communities (i.e., simulation replicates). Using the last `r unique(list_sim[[2]]$n_timestep)` time steps, we obtained the following summary statistics of the total community density $\sum_i^SN_i$ to examine the community-level response to intentional release: coefficient of variation (CV), number of species persist (defined as $N_i > 0.01$ for the entire simulation), temporal mean ($\mu$), and temporal SD ($\sigma$). We also calculated the temporal mean and SD for the enhanced ($N_1$) and unenhanced species ($\sum_{i,i \ne 1}^S N_i$) separately to infer underlying mechanisms.

When carrying capacity was small ($K=$ `r min(list_param[[2]]$k`), our model predicted a destabilizing effect of intentional release on ecological communities as illustrated by increased CV with increasing numbers of releases (**Figure 1A**). This pattern stemmed mainly from the reduced mean of the total community density, and both enhanced and unenhanced species groups were responsible (**Figure 1A**). The enhanced species decreased because release induced the negative competitive effect exceeding the reproductive contribution of released individuals [@satakeStockingCaptivebredFish2012]. Meanwhile, interspecific competition reduced the unenhanced species at high levels of intentional release, resulting in fewer persisting species (**Figure 1A**). Combined, the total community density decreased more sharply than individual species groups (**Figure 1A**). The SDs showed a similar trend, but the relationship was flatter at the community level (**Figure 1A**). Since a CV is a ratio of an SD to a mean, the steeper decline of the mean community density led to the increased CV. These patterns were qualitatively similar across most ecological contexts with the low carrying capacity (robust to the changes in $r_1$, $\bar{\alpha}$, and $f_R$; **Figures S1-S3**).

@@

Intentional release, however, had little or a stabilizing influence on community dynamics when carrying capacity was sufficiently large ($K=$ `r max(list_param[[2]]$k`); **Figures S4-S6**). In particular, intentional release increased the enhanced species with a low population growth rate ($r_1=$ `r min(list_param[[2]]$r1)`; **Figure S4**). This pattern suggests that intentional release may be effective for species or in environments with limited intraspecific competition (= $r_1 K^{-1}$) and is consistent with the fact that some of the best evidence for successful intentional release comes from re-introduction/translocation programs of extirpated species or population enhancement of long-lived endangered species. In the Sado island (Japan), for example, the endangered Crested Ibis *Nipponia nippon* showed exponential growth of the population since the initial reintroduction of ten captive-bred individuals in 2008. Similarly, translocations of Humpback chub, a federally-listed fish species, seem highly cost-effective in enhancing their tributary populations in the Little Colorado River. It is important to note that there are two common features in these successful examples. First, the receiving ecosystem remains nearly-intact (i.e., high carrying capacity). Second, the number of releases is rather small unlike release programs for commercial and recreational species. These conditions meet the theoretical assumptions with which intentional release led to increased abundance of the enhanced species with negligible impacts on other species. However, the preponderance of unsuccessful reintroduction programs implies that these conditions are not always sufficed. Our analysis suggests that pre-release examination and restoration of environmental capacity may be a key to successful release programs with minimal impacts on other community members.

>>>>

Human demands for natural resources are ever-increasing, such that active interventions are critical to the sustainable management of fisheries, forestry, and wildlife [@laikreCompromisingGeneticDiversity2010]. Releases of captive-bred native species ("intentional release") is a form of the efforts to enhance wild populations of diverse taxa [@snyderLimitationsCaptiveBreeding1996; @laikreCompromisingGeneticDiversity2010]. The magnitude of intentional release for commercial and recreational species is massive; typically, a single program releases tens of thousands to millions of plant, invertebrate (bivalves @haagNorthAmericanFreshwater2012, insects), and vertebrate individuals (fish, birds) into the wild annually (reviewed in [@laikreCompromisingGeneticDiversity2010; @kitadaEconomicEcologicalGenetic2018]). There is growing awareness that intentional release entails ecological risks, including the accumulation of deleterious alleles [@arakiGeneticEffectsCaptive2007], the spread of diseases and parasites in wild populations [@krkosekDecliningWildSalmon2007], and the intensification of intraspecific competition within the enhanced species. Yet, this method is still pervasive in natural resource management likely because of the apparent economic benefit of release programs. For example, Kitada @kitadaEconomicEcologicalGenetic2018 analyzed economic performance of release programs for 12 major fishery resources -- the gross monetary yield exceeded the seed-production cost in nine species, suggesting that release programs are economically profitable in general.

Current debates, however, overlook the fact that we have rarely assessed the community-wide impact of intentional release. Species are all embedded in the complex web of interacting organisms, and the stable coexistence of competing species through density-dependent feedback underpins the emergent stability of ecological communities [@tilmanEcologicalConsequencesChanges1999; @carrollNicheFitnessDifferences2011a; @loreauDoesFunctionalRedundancy2004]. Intentional release may disrupt the sensitive balance of species interactions because it introduces unnaturally high numbers of individuals into the wild [@laikreCompromisingGeneticDiversity2010; @careyNativeInvadersChallenges2012; @kitadaEconomicEcologicalGenetic2018]. Hence, this form of species management may intervene in the ecological process that allows competing species to coexist, ultimately degrading the long-term community stability. The community-wide impact must be particularly serious in release programs for fishery resources as many of them release millions of hatchery individuals into a single water body (e.g., a watershed) every year, which is likely far beyond the environmental capacity of the recipient ecosystem. Evidence for this hypothesis is lacking, however.

@@

Prior to our full-community analysis, we analyzed a two species community to understand the model behavior. The prior analysis revealed distinct parameter spaces predicting destabilizing, neutral, and stabilizing effects of intentional release on community dynamics. In general, intentional release had a neutral to destabilizing effect in a community with slow-growing species ($r_i < 1.5$). Meanwhile, for higher values of $r_i$ ($\ge 1.5$), intentional release stabilized community dynamics by dampening limit cycles or chaos. Carrying capacity and competition also played an important role in shifting the parameter spaces. According to this result, we chose `r nrow(list_param[[2]])` simulation scenarios, differing in the intrinsic growth of an enhanced species $r_1$, competition $\bar{\alpha}$, carrying capacity $K$, and relative fitness $f_R$ (see **Methods**). Under each parameter scenario, we ran `r attr(list_param[[2]], "n_sim")` time steps of `r max(list_sim[[2]]$n_rep)` independent communities (i.e., simulation replicates). Using the last `r unique(list_sim[[2]]$n_timestep)` time steps, we obtained the following summary statistics of the total community density $\sum_i^SN_i$ to examine the community-level response to intentional release: coefficient of variation (CV), number of species persist (defined as $N_i > 0.01$ for the entire simulation), temporal mean ($\mu$), and temporal SD ($\sigma$). We also calculated the temporal mean and SD for the enhanced ($N_1$) and unenhanced species ($\sum_{i,i \ne 1}^S N_i$) separately to infer underlying mechanisms.

When carrying capacity was small ($K=$ `r min(list_param[[2]]$k)`), our model predicted a destabilizing effect of intentional release on ecological communities as illustrated by increased CV with increasing numbers of releases (**Figure 1A**). This pattern stemmed mainly from the reduced mean of the total community density, and both enhanced and unenhanced species groups were responsible (**Figure 1A**). The enhanced species decreased because release induced the negative competitive effect exceeding the reproductive contribution of released individuals [@satakeStockingCaptivebredFish2012]. Meanwhile, interspecific competition reduced the unenhanced species at high levels of intentional release, resulting in fewer persisting species (**Figure 1A**). Combined, the total community density decreased more sharply than individual species groups (**Figure 1A**). The SDs showed a similar trend, but the relationship was flatter at the community level (**Figure 1A**). Since a CV is a ratio of an SD to a mean, the steeper decline of the mean community density led to the increased CV. These patterns were qualitatively similar across ecological contexts with the low carrying capacity (robust to the changes in $r_1$, $\bar{\alpha}$, and $f_R$; **Figures S1-S3**).

@@

Intentional release, however, had little or a stabilizing influence on community dynamics when carrying capacity was sufficiently large ($K=$ `r max(list_param[[2]]$k)`; **Figures S4-S6**). In particular, intentional release increased the enhanced species with a low population growth rate ($r_1=$ `r min(list_param[[2]]$r1)`; **Figure S4**), suggesting that weak intraspecific competition within the enhanced species (= $r_1 K^{-1}$) is needed for intentional release to be effective. Our result is reasonable because some of the best evidence for successful intentional release comes from re-introduction/translocation programs of extirpated species or population enhancement of long-lived endangered species. In the Sado island (Japan), for example, the endangered Crested Ibis *Nipponia nippon* showed exponential growth of the population since the initial reintroduction of ten captive-bred individuals in 2008. Similarly, translocations of Humpback chub, a federally-listed fish species, seem highly cost-effective in enhancing their tributary populations in the Little Colorado River. Two common features exist in these successful examples. First, the receiving ecosystem remains nearly-intact (i.e., high carrying capacity). Second, the number of releases in these conservation-oriented programs is rather small (several tens to hundreds) unlike release programs for commercial and recreational species (up to several tens of millions). These conditions meet the theoretical assumptions with which intentional release led to increased abundance of the enhanced species with negligible impacts on other species. However, the preponderance of unsuccessful reintroduction programs implies that these conditions are not always sufficed. In several examples, excessive numbers of individuals are released into degraded ecosystems with absolute failures of population enhancement or establishment. Our analysis suggests that pre-release examination and restoration of environmental capacity may be a key to successful release programs with minimal impacts on other community members.
